Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-15 paper/paper_maddie_edits.tex
added remaining comments and questions

Figure~\ref{fig:benchmark}(a) shows the time and space complexity of tensor network contraction for different graph sizes, where the space complexity is the same as the treewidth of the problem graph.
The contraction order is obtained using the local search algorithm in Ref.~\cite{Kalachev2021}.
In practice, slicing technique is used for graphs with treewidth greater than 27 to fit the computation into a 16GB memory.
One can see that all the computing times in figure~\ref{fig:benchmark}(b), (c) and (d) have a strong correlation with the treewidth.
Among these benchmarks, computational tasks with data types \texttt{T (CPU)},
\texttt{$\mathbbm{R}$ (CPU)}, \texttt{$\mathbbm{R}$ (GPU)}, \texttt{$\mathbbm{C}$ (CPU)}, \texttt{$\mathbbm{C}$ (GPU)} and \texttt{T+bounding (CPU)} can utilize fast BLAS functions to evaluate tensor contractions, hence 
are much faster comparing to non-BLAS element types in the same category.
GPU computes much faster than CPU in all cases when the problem scale is large enough such that the actual computing time is comparable or larger than the launching overhead of CUDA kernels.
Most algebras can be computed on a GPU, except those requiring dynamic sized structures, i.e.\ \texttt{PN}, \texttt{SN}, \texttt{T100} and \texttt{SPT}.
In figure~\ref{fig:benchmark}(c), one can see the Fourier transformation based method is the fastest in computing the independence polynomial,
but it may suffer from round-off errors. The finite field (GF$(p)$) approach is the only method that does not have round-off errors and can be run on a GPU.
In figure~\ref{fig:benchmark}(d), one can see the technique to bound the enumeration space improves the performance for more than one order of magnitude in enumerating the MISs.
Bounding can also reduce the memory usage significantly, without which the largest computable graph size is only $\sim150$ on a device with 32GB main memory.

@@

A useful number that can be computed using the number of independent sets is the entropy constant for the hardcore lattice gases on these graphs. For the square grid graphs, it is called the \textit{hard square entropy constant} (\href{https://oeis.org/A085850}{OEIS A085850}), which is defined as $\lim_{L\rightarrow \infty} F(L,L)^{1/L^2}$, where $F(L,L)$ is the number of independent sets of a given lattice dimensions $L \times L$. This quantity arises in statistical mechanics of hard-square lattice gases~\cite{Baxter1980, Pearce1988} and is used to understand phase transitions for these systems. This entropy constant is not known to have an exact representation, but it is accurately known in many digits. Similarly, we can define entropy constants for other lattice gases. In \Fig{fig:hardsquare}, we look at how $F(L,L)^{1/\lfloor pL^2\rceil}$ scales as a function of the grid size $L$ for all types of graphs shown in Figure~\ref{fig:lattices}. Our results match the known results for the non-disordered square grid and King's graphs. For disordered square grid and King's graphs with a filling factor $p=0.8$, we randomly sample 1000 graph instances. To our knowledge, the entropy constants for these disordered graphs have not been studied before. Interestingly, the variations due to different random instances are negligible for this entropy quantity.

@@

\subsection{The overlap gap property}
With the tool to enumerate configurations, one can try to understand the structure of the independent set configuration space, such as the optimization landscape for finding the MISs and the geometry of the solution space. One of the known barriers for finding the MIS is the so-called overlap gap property~\cite{Gamarnik2013, Gamarnik2019}. Intuitively, if the overlap gap property is present, it means every two large independent sets either have a significant intersection or very small intersection; it implies that large independent sets are clustered together. This clustering property has been used to prove the limitations of local algorithms in finding the MISs~\cite{Gamarnik2013, Gamarnik2019}. Here, we use the computed configurations of large independent sets to investigate the presence or absence of the overlap gap property for some given graphs. In particular, in \Fig{fig:hamming}, we enumerated all ISs of size $\geq\alpha(G)-1$ for two random three regular graph instances of size 100 and show the pairwise Hamming distance statistics for the enumerated configurations. In \Fig{fig:hamming}(a), we observed the multiple peak structure for the pairwise Hamming distance for one of the graphs. This indicates the presence of the overlap gap property and disconnected clusters exist in the configuration space for large independent sets. We expect our numerical tool can be used to understand this phenomenon better and to further investigate the graph properties and the geometry of the configuration spaces for a variety of graph instances.
\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig3.pdf}
    \caption{The distribution of pairwise Hamming distances between every two independent sets of size $\geq\alpha(G)-1$ for two random three regular graph instances of size $100$.}
    \label{fig:hamming}
\end{figure}

@@

    \caption{The the tensor network element types used in this work, while the purpose of these element types can be found in \Tbl{tbl:generictypes}.
    The overlap between two algebra indicates that a new algebra can be created by combining those two types of algebra.}
     \label{fig:venn-diagram}
\end{figure}

>>>>

Figure~\ref{fig:benchmark}(a) shows the time and space complexity of tensor network contraction for different graph sizes, where the space complexity is the same as the treewidth of the problem graph.
The contraction order is obtained using the local search algorithm in Ref.~\cite{Kalachev2021}.
In practice, slicing technique is used for graphs with treewidth greater than 27 to fit the computation into a 16GB memory. \purple{What is the slicing technique? Can you add a refernce if it's from another paper?}
One can see that all the computing times in figure~\ref{fig:benchmark}(b), (c) and (d) have a strong correlation with the treewidth \purple{with the number of vertices, and therefore the treewidth}. \purple{Does this say anything about how optimized the contraction order is? If so, it might be worth explicitly saying. Some of the quantities in panel d look like they might be superexponential in the number of vertices (tree-width).}
Among these benchmarks, computational tasks with data types \texttt{T (CPU)},
\texttt{$\mathbbm{R}$ (CPU)}, \texttt{$\mathbbm{R}$ (GPU)}, \texttt{$\mathbbm{C}$ (CPU)}, \texttt{$\mathbbm{C}$ (GPU)} and \texttt{T+bounding (CPU)} can utilize fast BLAS functions to evaluate tensor contractions, hence 
are much faster comparing to non-BLAS element types in the same category. \purple{It might be worth explicitly saying the data types the first time they're mentioned (for example, I don't know what \texttt{T} is. Also, it might be worth adding a couple descriptor words on bounding.}
GPU computes much faster than CPU in all cases when the problem scale is large enough such that the actual computing time is comparable or larger than the launching overhead of CUDA kernels.
Most algebras can be computed on a GPU, except those requiring dynamic sized structures, i.e.\ \texttt{PN}, \texttt{SN}, \texttt{T100} and \texttt{SPT} \purple{What are these things listed?}.
In figure~\ref{fig:benchmark}(c), one can see the Fourier transformation based method is the fastest in computing the independence polynomial,
but it may suffer from round-off errors \purple{Can the round-off error thing be seen from Figure c?}. The finite field (GF$(p)$) approach is the only method that does not have round-off errors and can be run on a GPU.
In figure~\ref{fig:benchmark}(d), one can see the technique to bound the enumeration space improves the performance for more than one order of magnitude in enumerating the MISs.
Bounding can also reduce the memory usage significantly, without which the largest computable graph size is only $\sim150$ on a device with 32GB main memory.

@@

A useful number that can be computed using the number of independent sets is the entropy constant for the hardcore lattice gases on these graphs. \purple{Why is it useful and what does it tell us about the physical system?} For the square grid graphs, it is called the \textit{hard square entropy constant} (\href{https://oeis.org/A085850}{OEIS A085850}), which is defined as $\lim_{L\rightarrow \infty} F(L,L)^{1/L^2}$, where $F(L,L)$ is the number of independent sets of a given lattice dimensions $L \times L$. This quantity arises in statistical mechanics of hard-square lattice gases~\cite{Baxter1980, Pearce1988} and is used to understand phase transitions for these systems. This entropy constant is not known to have an exact representation, but it is accurately known in many digits. Similarly, we can define entropy constants for other lattice gases. In \Fig{fig:hardsquare}, we look at how $F(L,L)^{1/\lfloor pL^2\rceil}$ scales as a function of the grid size $L$ for all types of graphs shown in Figure~\ref{fig:lattices}. Our results match the known results for the non-disordered square grid and King's graphs. For disordered square grid and King's graphs with a filling factor $p=0.8$, we randomly sample 1000 graph instances. To our knowledge, the entropy constants for these disordered graphs have not been studied before. Interestingly, the variations due to different random instances are negligible for this entropy quantity.

@@

\subsection{The overlap gap property}
With the tool to enumerate configurations, one can try to understand the structure of the independent set configuration space, such as the optimization landscape for finding the MISs and the geometry of the solution space. One of the known barriers for finding the MIS is the so-called overlap gap property~\cite{Gamarnik2013, Gamarnik2019}. Intuitively, if the overlap gap property is present, it means every two large independent sets either have a significant intersection or very small intersection; it implies that large independent sets are clustered together. This clustering property has been used to prove rigorous bounds on the performance of local algorithms in finding the MISs~\cite{Gamarnik2013, Gamarnik2019}. Here, we use the computed configurations of large independent sets to investigate the presence or absence of the overlap gap property for some given graphs. In particular, in \Fig{fig:hamming}, we enumerated all ISs of size $\geq\alpha(G)-1$ for two random three regular graph instances of size 100 and show the pairwise Hamming distance statistics for the enumerated configurations. In \Fig{fig:hamming}(a), we observed the multiple peak structure for the pairwise Hamming distance for one of the graphs. This indicates the presence of the overlap gap property and disconnected clusters exist in the configuration space for large independent sets. We expect our numerical tool can be used to understand this phenomenon better and to further investigate the graph properties and the geometry of the configuration spaces for a variety of graph instances.
\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig3.pdf}
    \caption{The distribution of pairwise Hamming distances between every two independent sets of size $\geq\alpha(G)-1$ for two random three regular graph instances of size $100$.}
    \label{fig:hamming}
\end{figure}

@@

    \caption{The tensor network element types used in this work, while the purpose of these element types can be found in \Tbl{tbl:generictypes}.
    The overlap between two algebra indicates that a new algebra can be created by combining those two types of algebra. \purple{It's not clear what some words mean on this diagram, such as ``highest order'' and ``extended tropical''.}}
     \label{fig:venn-diagram}
\end{figure}
